A Note on Convergence Diagnostics for Multiple Imputation Algorithms

### 2019-12-14 Simulation/Writeup/NoteConvergence.Rnw
shorter abstract

%% - \Abstract{} almost as usual
\Abstract{
The ubiquitous problem of missing information may be circumvented by implementing the framework of multiple imputation (MI). With MI, missing data points are `filled in' (i.e. imputed) several times, and statistical inference is performed on each imputed data set. The validity of inferences may be threatened by non-convergence of this iterative algorithmic procedure, but diagnostics to evaluate algorithmic convergence have not been implemented in MI software. This note investigates the performance of two convergence diagnostics (\citep[$\widehat{R}$,][]{gelm92}; \citep[autocorrelation,][]{scha97}) through model-based simulation. To investigate convergence of the algorithm, the number of iterations is varied between one and one-hundred. Each of these 100 simulation conditions is replicated 1000 times. The convergence diagnostics are evaluated against three recommended simulation diagnostics (bias, confidence interval width, and coverage rate \citep{buur18}, since no benchmark diagnostic is available. This simulation study shows that three iterations may be sufficient to obtain ... \textbf{This Research Report contains a simulation study that serves as the basis of the technical paper that will be submitted for publication in \emph{Journal of Statistical Software}. I aim to publish a pre-print of this research report on \href{https://arxiv.org/}{arXiv}. That way, I can refer to the simulation study described here, without `bulking up' the technical paper that I want to submit for publication in JSS. Another option would be to attach this note as an online appendix to the technical paper on ShinyMICE.}  
% This Research Report contains a simulation study that serves as the basis of the technical paper that will be submitted for publication in \emph{Journal of Statistical Software}. I have chosen to use the first format from the Research Report guidelines: ``\emph{It is written as a (mini) thesis, with an introduction, methods section, some results (i.e., preliminary analyses, or pilot simulations), and a discussion of results. The length of the research report should be maximally 2500 words of text (without references list and or tables and figures). Please do not include appendices, and no more than 6 tables or figures. Table and Figure captions do not count towards the word limit. An abstract may be included, but is not necessary}''. I aim to publish a pre-print of this research report on \href{https://arxiv.org/}{ArXiv}. That way, I can refer to the simulation study described here, without 'bulking up' the technical paper that I want to submit for publication in JSS. Another option would be to attach this note as online appendix to the technical paper on ShinyMICE.
  }

@@

The intended audience of this note consists of empirical researchers and statisticians who use multiple imputation to solve missing data problems. Basic familiarity with MI methodology is assumed. For the theoretical foundation of MI, see \cite{rubin87}. For an accessible and comprehensive introduction to MI from an applied perspective, see \cite{buur18}.

The convergence guidelines introduced in this paper are developed to be integrated into the \pkg{mice} environment \citep{mice} in \proglang{R} \citep{R}. This note, therefore, follows notation and conventions of \cite{mice}. Deviations from the `original' notation by \cite{rubin87} are described in \citep[\S~2.2.3]{buur18}.

Let $Y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample. The collection of observed data values in $Y$ is denoted as $Y_{obs}$; the missing part of $Y$ is referred to as $Y_{mis}$. \textbf{REMOVE? Response indicator $R$ shows whether a data value in $Y$ is missing or observed.} The relation between $R$, $Y_{obs}$, and $Y_{mis}$ determines the missingness mechanism. This note only considers a `missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $Y$.

In this note, the terms `unobserved' and `missing' data are used interchangeably to refer to $Y_{mis}$. The terms `incomplete' or `observed' data denote $Y_{obs}$. Incomplete data is the starting point of the multiple imputation procedure. Figure \ref{fig:steps} provides an overview of the steps involved with MI.

@@

Missing data in $Y$ is `imputed' (i.e., filled in) $m$ times. The imputed data is combined with observed data $Y_{obs}$ to create $m$ completed data sets. On each completed data set, the analysis of scientific interest (or `complete data model') is performed. The quantity of scientific interest (e.g., a regression coefficient) is denoted with $Q$. Since $Q$ is estimated on each completed data set, $m$ separate $\hat{Q}$ values are obtained. These $m$ values are combined into a single pooled estimate $\bar{Q}$.

This note focuses on the algorithmic properties of the imputation step. The algorithm employed within this step has an iterative nature. That is, before drawing $m$ imputed values for each missing data point in $Y_{mis}$, a `chain' of potential values is sampled. Each of the $m$ chains starts with an initial value, drawn randomly from $Y_{obs}$. The chains are terminated after a predefined number of iterations. Only the ultimate sample that a chain lands on is imputed and subsequently used in the analysis and pooling steps. The collection of samples across iterations (between the initial value and the imputed value) for each of the $m$ imputations will be referred to as an `imputation chain'. The total number of iterations per chain will be denoted with $T$, where $t$ varies over the integers $1, 2, \dots, T$.

@@

We can interpret %the sign and magnitude of ... Any non-zero $AC$-value ($-1 \leq AC \leq 1$) . $AC = 0$ indicates independence between $\theta$-values.
$AC$-values as a measure of dependence within chains. Positive $AC$-values indicate recurrence (i.e., similarity between $\theta$-values of subsequent iterations), which may lead to trending. Negative $AC$-values occur when the $\theta$-values of subsequent iterations diverge from oneanother. Dependence in this sense does not threaten the stationarity component of convergence. %that we aim to evaluate with $AC$. 
On the contrary even---it may increase the variance of $\theta$. \textbf{speed up convergence???}  %have also indicate dependence, but do not threaten the stationarity component of convergence \textbf{[REMOVE INTERPRETATION HERE?]}. Instead of recurrence, there would be a `seesaw' effect within chains: subsequent $\theta$-values would deviate in opposite directions from the chain average $\theta^{(\cdot m)})$, increasing the variance of $\theta$. 
As convergence diagnostic, the interest is in positive $AC$-values.%\footnotemark 
\footnote{The magnitude of $AC$-values can also be evaluated statistically, but that is outside of this note's scope.}

@@

%% --------------------------------
\subsection{Convergence diagnostics}
It is apparant that there is a relation between convergence diagnostics $\widehat{R}$ and $AC$, and the number of iterations per simulation condition($T$). Generally speaking, we see a trend towards convergence as $T$ increases with each simulation condition ($\widehat{R}$-values approache one, and $AC$-values approache zero). The convergence diagnostics show more or less equivalent trends for chain means versus chain variances. We, therefore, only discuss the $\widehat{R}$ and $AC$-values of chain means. %convergence diagnostics for the chain means. %do not differentiate in the interpretation of chain means .

Figure \ref{fig:conv}A shows that $\widehat{R}$-values generally decrease with increasing imputation chain lengths. The decline stabilizes somewhere between the simulation conditions $T=30$ and $T=50$. The downward trend is most pronounced for $T=3$, and between $T = 5$ and $T = 10$. In the intervening conditions ($3 \leq T \leq 5$), however, we observe a steep increase in $\widehat{R}$-values. This increase implies that convergence cannot be diagnosed at $T=3$, as would be the case based on the conventional threshold $\widehat{R} < 1.2$. \textbf{This may be spurious.} According to the widely used threshold $\widehat{R} < 1.1$, convergence can be diagnosed for conditions where $T>9$. If we use the recently recommended threshold $\widehat{R} < 1.01$, we would conclude that we cannot diagnose convergence in any simulation condition. \textbf{add something about Rh below 1.}

@@

%% --------------------------------
\subsection{Simulation diagnostics}
We use average bias, average condidence interval width, and coverage rate as performance measures to evaluate $\widehat{R}$ and $AC$. We make the general observation that the simulation diagnostics behave as theorized for most simulation conditions (bias around zero, coverage rate around 95\%).

Figure \ref{fig:sim}A shows that bias is fairly stable across simulation conditions. The condition $T=1$ clearly deviates from this trend with a negative bias. The bias for $T=2$ is below average, but within the range of fluctuations. Conditions where $T>3$ \textbf{are ok, Loess line is flat for $T>20$}.  
%the average bias between the regression coefficient of $X$ on $Y$ in the complete dataset, and $\bar{Q}$, \textbf{the pooled estimated regression coefficient and the ... Average bias is close to zero for simulation conditions with two iterations or more. The simulation condition with one iteration has a negative bias. From two iterations upwards, the average bias across repetitions is stable. Across iterations two to one hundred, bias fluctuates within a narrow range around zero. Loess line is flat around it = 20.} 
% Average bias is not zero because there is only one incomplete dataset to be imputed. Bias is a sample effect.
%Figure \ref{fig:sim}A shows that bias is fairly stable across simulation conditions. The condition $T=1$ clearly deviates from this trend with a negative bias. The bias for $T=2$ is below average, but within the range of fluctuations.  Conditions where $T>3$ \textbf{are ok, Loess line is flat for $T>20$}.

@@

\textbf{From the simulation diagnostics, we see that simulation conditions where $T>3$ are generally ok. The performance measures indicate that for the current data, MI procedure, and quantity of interest, the algorithm should have maxit argument of four or more. This is, however, a simple simulation problem, with MCAR missingness mechanism, univariate missingness, and moderate fmi's [REMOVE?? COMPUTE??]. harder problems require more iterations.} %give some implications for these measures. connect convergence and simulation diagnostics.}

There is a discrepancy between what the convergence diagnostics and the performance measures indicate. While $T=4$ is ok with repsect to simulation quantities, $T=3$ is the worst possible value according to the convergence diagnostics, and $T=4$ provides marginally better convergence diagnostics.

>>>>

%% - \Abstract{} almost as usual
\Abstract{
This note investigates how algorithmic convergence of multiple imputation procedures---methodology to circumvent the ubiquitous problem of missing data---may be diagnosed. We conclude that potential scale reduction factor $\widehat{R}$ and autocorrelation may be appropriate convergence diagnostics, but their performance differs from conventional applications to iterative algorithmic procedures. These results will be implemented into an interactive framework for the evaluation of multiple imputation methods that is under development as research Master thesis project.
%%%%%%%%%%%%%
%The ubiquitous problem of missing information may be circumvented by implementing the framework of multiple imputation (MI). With MI, missing data points are `filled in' (i.e. imputed) several times, and statistical inference is performed on each imputed data set. The validity of inferences may be threatened by non-convergence of this iterative algorithmic procedure, but diagnostics to evaluate algorithmic convergence have not been implemented in MI software. This note investigates the performance of two convergence diagnostics (\citep[$\widehat{R}$,][]{gelm92}; \citep[autocorrelation,][]{scha97}) through model-based simulation. To investigate convergence of the algorithm, the number of iterations is varied between one and one-hundred. Each of these 100 simulation conditions is replicated 1000 times. The convergence diagnostics are evaluated against three recommended simulation diagnostics (bias, confidence interval width, and coverage rate \citep{buur18}, since no benchmark diagnostic is available. This simulation study shows that three iterations may be sufficient to obtain ... \textbf{This Research Report contains a simulation study that serves as the basis of the technical paper that will be submitted for publication in \emph{Journal of Statistical Software}. I aim to publish a pre-print of this research report on \href{https://arxiv.org/}{arXiv}. That way, I can refer to the simulation study described here, without `bulking up' the technical paper that I want to submit for publication in JSS. Another option would be to attach this note as an online appendix to the technical paper on ShinyMICE.}  
%%%%%%%%%%%%
% This Research Report contains a simulation study that serves as the basis of the technical paper that will be submitted for publication in \emph{Journal of Statistical Software}. I have chosen to use the first format from the Research Report guidelines: ``\emph{It is written as a (mini) thesis, with an introduction, methods section, some results (i.e., preliminary analyses, or pilot simulations), and a discussion of results. The length of the research report should be maximally 2500 words of text (without references list and or tables and figures). Please do not include appendices, and no more than 6 tables or figures. Table and Figure captions do not count towards the word limit. An abstract may be included, but is not necessary}''. I aim to publish a pre-print of this research report on \href{https://arxiv.org/}{ArXiv}. That way, I can refer to the simulation study described here, without 'bulking up' the technical paper that I want to submit for publication in JSS. Another option would be to attach this note as online appendix to the technical paper on ShinyMICE.
  }

@@

The intended audience of this note consists of anyone who uses multiple imputation to solve missing data problems. Basic familiarity with MI methodology is assumed.\footnote{For the theoretical foundation of MI, see \cite{rubin87}. For an accessible and comprehensive introduction to MI from an applied perspective, see \cite{buur18}.}  
This note follows notation and conventions of \cite{mice}. % to suit the \pkg{mice} environment. %Deviations from the `original' notation by \cite{rubin87} are described in \citep[\S~2.2.3]{buur18}.

Let $Y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample. The collection of observed data values in $Y$ is denoted by $Y_{obs}$, and will be referred to as `incomplete' or `observed' data interchangeably. The missing part of $Y$ is denoted by $Y_{mis}$. Which parts of $Y$ are missing is determined by the `missingness mechanism'. %The relation between $Y_{obs}$ and $Y_{mis}$ \textbf{REMOVE? Response indicator $R$ shows whether a data value in $Y$ is missing or observed.} The relation between $R$, $Y_{obs}$, and $Y_{mis}$ determines the missingness mechanism. 
This note only considers a `missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $Y$.

%In this note, the terms `unobserved' and `missing' data are used interchangeably to refer to $Y_{mis}$. The terms `incomplete' or `observed' data denote $Y_{obs}$. Incomplete data is the starting point of the multiple imputation procedure.

Figure \ref{fig:steps} provides an overview of the steps involved with MI. Missing data in $Y$ is `imputed' (i.e., filled in) $m$ times. The imputed data is combined with observed data $Y_{obs}$ to create $m$ completed data sets. On each completed data set, the analysis of scientific interest (or `complete data model') is performed. The quantity of scientific interest (e.g., a regression coefficient) is denoted with $Q$. Since $Q$ is estimated on each completed data set, $m$ separate $\hat{Q}$ values are obtained. These $m$ values are combined into a single pooled estimate $\bar{Q}$.

@@

This note focuses on the algorithmic properties of the imputation step. The algorithm employed within this step has an iterative nature. That is, before drawing $m$ imputed values for each missing data point in $Y_{mis}$, a `chain' of potential values is sampled. Each of the $m$ chains starts with an initial value, drawn randomly from $Y_{obs}$. The chains are terminated after a predefined number of iterations. Only the ultimate sample that a chain lands on is imputed, and subsequently used in the analysis and pooling steps. The collection of samples across iterations (between the initial value and the imputed value) for each of the $m$ imputations will be referred to as an `imputation chain'. The total number of iterations per chain will be denoted with $T$, where $t$ varies over the integers $1, 2, \dots, T$.

@@

We can interpret %the sign and magnitude of ... Any non-zero $AC$-value ($-1 \leq AC \leq 1$) . $AC = 0$ indicates independence between $\theta$-values.
$AC$-values as a measure of dependence within chains. Positive $AC$-values indicate recurrence (i.e., similarity between $\theta$-values of subsequent iterations), which may lead to trending. Negative $AC$-values occur when the $\theta$-values of subsequent iterations diverge from one-another. Dependence in this sense does not threaten the stationarity component of convergence. %that we aim to evaluate with $AC$. 
On the contrary even---it may increase the variance of $\theta$. \textbf{speed up convergence???}  %have also indicate dependence, but do not threaten the stationarity component of convergence \textbf{[REMOVE INTERPRETATION HERE?]}. Instead of recurrence, there would be a `seesaw' effect within chains: subsequent $\theta$-values would deviate in opposite directions from the chain average $\theta^{(\cdot m)})$, increasing the variance of $\theta$. 
As convergence diagnostic, the interest is in positive $AC$-values.%\footnotemark 
\footnote{The magnitude of $AC$-values can also be evaluated statistically, but that is outside of this note's scope.}

@@

%% --------------------------------
\subsection{Convergence diagnostics}
It is apparent that there is a relation between convergence diagnostics $\widehat{R}$ and $AC$, and the number of iterations per simulation condition($T$). Generally speaking, we see a trend towards convergence as $T$ increases with each simulation condition ($\widehat{R}$-values approach one, and $AC$-values approach zero). The convergence diagnostics show more or less equivalent trends for chain means versus chain variances. We, therefore, only discuss the $\widehat{R}$ and $AC$-values of chain means. %convergence diagnostics for the chain means. %do not differentiate in the interpretation of chain means .

Figure \ref{fig:conv}A shows that $\widehat{R}$-values generally decrease with increasing imputation chain lengths. The decline stabilizes somewhere between the simulation conditions $T=30$ and $T=50$. The downward trend is most pronounced for $T=3$, and between $T = 5$ and $T = 10$. In the intervening conditions ($3 \leq T \leq 5$), however, we observe a steep increase in $\widehat{R}$-values. This increase implies that convergence cannot be diagnosed at $T=3$, as would be the case based on the conventional threshold $\widehat{R} < 1.2$. \textbf{[This may be spurious??]} According to the widely used threshold $\widehat{R} < 1.1$, convergence can be diagnosed for conditions where $T>9$. If we use the recently recommended threshold $\widehat{R} < 1.01$, we would conclude that we cannot diagnose convergence in any simulation condition. \textbf{[Add something about $\widehat{R} <1$?]}

@@

%% --------------------------------
\subsection{Simulation diagnostics}
We use average bias, average confidence interval width, and coverage rate as performance measures to evaluate $\widehat{R}$ and $AC$. We make the general observation that the simulation diagnostics behave as theorized for most simulation conditions (bias around zero, coverage rate around 95\%).

Figure \ref{fig:sim}A shows that bias is fairly stable across simulation conditions. The condition $T=1$ clearly deviates from this trend with a negative bias. The bias for $T=2$ is below average, but within the range of fluctuations. Conditions where $T>3$ \textbf{are OK, Loess line is flat for $T>20$}.  
%the average bias between the regression coefficient of $X$ on $Y$ in the complete dataset, and $\bar{Q}$, \textbf{the pooled estimated regression coefficient and the ... Average bias is close to zero for simulation conditions with two iterations or more. The simulation condition with one iteration has a negative bias. From two iterations upwards, the average bias across repetitions is stable. Across iterations two to one hundred, bias fluctuates within a narrow range around zero. Loess line is flat around it = 20.} 
% Average bias is not zero because there is only one incomplete dataset to be imputed. Bias is a sample effect.
%Figure \ref{fig:sim}A shows that bias is fairly stable across simulation conditions. The condition $T=1$ clearly deviates from this trend with a negative bias. The bias for $T=2$ is below average, but within the range of fluctuations.  Conditions where $T>3$ \textbf{are ok, Loess line is flat for $T>20$}.

@@

\textbf{From the simulation diagnostics, we see that simulation conditions where $T>3$ are generally OK. The performance measures indicate that for the current data, MI procedure, and quantity of interest, the algorithm should have maxit argument of four or more. This is, however, a simple simulation problem, with MCAR missingness mechanism, univariate missingness, and moderate fmi's [REMOVE?? COMPUTE??]. harder problems require more iterations.} %give some implications for these measures. connect convergence and simulation diagnostics.}

There is a discrepancy between what the convergence diagnostics and the performance measures indicate. While $T=4$ is OK with respect to simulation quantities, $T=3$ is the worst possible value according to the convergence diagnostics, and $T=4$ provides marginally better convergence diagnostics.
